---
pagetitle: "openwashdata"
page-layout: custom
section-divs: false
toc: false
link-external-icon: false
link-external-newwindow: false
execute:
  echo: false
  warning: false
  message: false
listing:
  - id: latest-posts
    contents: pages/blog/posts/*/index.qmd
    type: default
    max-items: 3
    sort: "date desc"
    fields: [date, title, author]
    sort-ui: false
    filter-ui: false
    categories: false
---

```{r}
#| include: false
library(dplyr)
# Same source as the data catalogue: the weekly owdata harvest of the
# organisation (openwashdata/owdata#11).
data <- readr::read_csv(
  "https://raw.githubusercontent.com/openwashdata/owdata/main/inst/extdata/owd_packages.csv",
  show_col_types = FALSE
)
n_pub <- nrow(data)
n_ds <- sum(data$n_datasets, na.rm = TRUE)
first_year <- min(as.integer(format(data$date, "%Y")), na.rm = TRUE)
n_years <- as.integer(format(Sys.Date(), "%Y")) - first_year + 1
```

::: {.owd-hero-a}
::: {.owd-section}

# Open data for water, sanitation and hygiene

<p class="owd-lead">openwashdata is an active global community that applies the FAIR principles to data from the water, sanitation and hygiene (WASH) sector, and teaches the skills to do it.</p>

::: {.owd-actions}
[Explore the data](pages/gallery/data/index.qmd){.btn .btn-primary}
[Read our stories](pages/blog/index.qmd){.btn .btn-outline}
:::

::: {#subscribe}
<form action="https://buttondown.com/api/emails/embed-subscribe/openwashdata" method="post" class="owd-subscribe">
<label for="owd-email-hero">One email a month: a dataset, a story, and the community's news.</label>
<input type="email" name="email" id="owd-email-hero" placeholder="your email address" required>
<input type="submit" value="Subscribe">
</form>
:::

:::
:::

::: {.owd-tint}
::: {.owd-section}
::: {.owd-numbers}
::: {.owd-number-item}
<div class="owd-number">`r n_pub`</div>
<div class="owd-number-label">open data packages</div>
:::
::: {.owd-number-item}
<div class="owd-number">`r n_ds`</div>
<div class="owd-number-label">datasets inside them</div>
:::
::: {.owd-number-item}
<div class="owd-number">`r n_years`</div>
<div class="owd-number-label">years of publishing, since `r first_year`</div>
:::
:::
:::
:::

::: {.owd-section}

## Dataset of the month

::: {.owd-dotm}
::: {.owd-dotm-text}
**gwinfraassess** holds an assessment of 123 groundwater monitoring wells in Malawi by BASEflow Malawi in 2022 and 2023. Field teams recorded the condition of each well, its water level logger, the measuring equipment and what the well needs to keep working. Every well but one has an improvement plan, and the action named most often is a trained person to read the gauge (76 wells), ahead of a manual dipper (74 wells).

[Open gwinfraassess](https://openwashdata.github.io/gwinfraassess){.owd-more}
:::

::: {.owd-dotm-chart}

<p class="owd-chart-title">The most common fix is a person</p>
<p class="owd-chart-note">Share of the 122 wells with an improvement plan that need each action, coloured by kind of action. <span class="owd-chart-hint">Hover over a bar for the number of wells.</span></p>

```{r}
#| label: dataset-of-the-month-data
#| echo: false
# pak::pak("openwashdata/gwinfraassess")
library(gwinfraassess)
library(dplyr)
library(tidyr)

action_groups <- c(
  "Identify and training gauge reader" = "People",
  "Provide manual dipper" = "Equipment",
  "Replacement of the level logger" = "Equipment",
  "Installation of a new logger" = "Equipment",
  "Replace barologger" = "Equipment",
  "Rehabilitation of the civil works" = "Structure",
  "Replace end cap" = "Structure",
  "Replacement of the cover and hinges" = "Structure",
  "Welding on the existing hinges" = "Structure",
  "Replace and weld the hinges" = "Structure",
  "Redevelop the monitoring well" = "Structure",
  "Replace the well" = "Replace or retire",
  "Discard monitoring well" = "Replace or retire",
  "Discard as monitoring well and donate to the community" = "Replace or retire"
)

action_labels <- c(
  "Identify and training gauge reader" = "Find and train a gauge reader",
  "Provide manual dipper" = "Provide a manual dipper",
  "Replacement of the level logger" = "Replace the level logger",
  "Installation of a new logger" = "Install a new logger",
  "Replace barologger" = "Replace the barologger",
  "Rehabilitation of the civil works" = "Repair the civil works",
  "Replace end cap" = "Replace the end cap",
  "Replacement of the cover and hinges" = "Replace cover and hinges",
  "Welding on the existing hinges" = "Weld the existing hinges",
  "Replace and weld the hinges" = "Replace and weld the hinges",
  "Redevelop the monitoring well" = "Redevelop the well",
  "Replace the well" = "Replace the well",
  "Discard monitoring well" = "Discard the well",
  "Discard as monitoring well and donate to the community" = "Discard, give to community"
)

wells <- gwinfraassess |>
  mutate(well = row_number()) |>
  filter(!is.na(improvement_plan))

n_plans <- nrow(wells)

actions <- wells |>
  separate_longer_delim(improvement_plan, ",") |>
  mutate(action = gsub("\\s+", " ", trimws(improvement_plan))) |>
  filter(action %in% names(action_groups)) |>
  distinct(well, action) |>
  count(action, name = "n") |>
  mutate(
    group = unname(action_groups[action]),
    label = unname(action_labels[action]),
    share = round(100 * n / n_plans)
  ) |>
  arrange(desc(n))

ojs_define(actions_data = actions, n_plans = n_plans)
```

```{ojs}
//| echo: false
actions = transpose(actions_data)

// width of the chart column, so the text keeps its size on every screen
chartWidth = Generators.observe((notify) => {
  const el = document.querySelector(".owd-dotm-chart");
  const report = () => notify(Math.max(280, el.clientWidth));
  const observer = new ResizeObserver(report);
  observer.observe(el);
  report();
  return () => observer.disconnect();
})

// chart-1, chart-3, chart-2 and a neutral grey from openwashdata/brand
palette = ["#8a3a8a", "#CF9E40", "#b04208", "#b9b9b9"]

Plot.plot({
  width: chartWidth,
  height: 400,
  marginLeft: chartWidth < 420 ? 170 : 200,
  marginRight: 36,
  marginTop: 4,
  marginBottom: 36,
  style: {fontFamily: "Atkinson Hyperlegible Next, sans-serif", fontSize: 13, background: "transparent"},
  ariaLabel: "Bar chart of improvement actions recommended for groundwater monitoring wells in Malawi",
  ariaDescription: "62 percent of wells need a trained gauge reader, 61 percent a manual dipper, 46 percent a replaced level logger and 42 percent repaired civil works.",
  x: {label: "% of wells", labelAnchor: "center", domain: [0, 70], grid: true, ticks: 4},
  y: {label: null, domain: actions.map((d) => d.label)},
  color: {
    domain: ["People", "Equipment", "Structure", "Replace or retire"],
    range: palette,
    legend: true
  },
  marks: [
    Plot.barX(actions, {
      y: "label",
      x: "share",
      fill: "group",
      tip: true,
      title: (d) => `${d.label}\n${d.n} of ${n_plans} wells (${d.share}%)\n${d.group}`
    }),
    Plot.text(actions, {y: "label", x: "share", text: (d) => `${d.share}%`, textAnchor: "start", dx: 4, fill: "#1e1e1e"}),
    Plot.ruleX([0])
  ]
})
```

:::
:::

:::

::: {.owd-tint}
::: {.owd-section}

## Learn

<p class="owd-lead">Data Science for openwashdata (ds4owd), our free online course, teaches WASH professionals data science in R. Two cohorts have graduated so far. A third cohort will start in September 2027. We are currently preparing the sign-up forms for the third iteration. In the meantime, you can browse the course website of the second cohort from 2025.</p>

[Course website of the 2025 cohort](https://ds4owd-002.github.io/website/){.owd-more}
[Projects of the 2026 graduates](pages/academy/graduates-2026/index.qmd){.owd-more}
[About the course](pages/academy/index.qmd){.owd-more}

:::
:::

::: {.owd-section}

## Read

<p class="owd-lead">Stories, tutorials and opinions from and for the community.</p>

::: {#latest-posts}
:::

[All posts](pages/blog/index.qmd){.owd-more}

:::

::: {.owd-tint}
::: {.owd-section}

## Join

```{=html}
<div class="owd-cards">
<a class="owd-card" href="pages/get-started/chat/index.qmd"><span class="owd-card-kicker">Talk</span><strong>Chat with us</strong><p>The community room on Matrix, open to everyone.</p></a>
<a class="owd-card" href="pages/blog/posts/2024-05-17-data-publication-1/index.qmd"><span class="owd-card-kicker">Share</span><strong>Publish your data</strong><p>We turn your WASH dataset into a documented, citable data package with you.</p></a>
<a class="owd-card" href="https://github.com/openwashdata"><span class="owd-card-kicker">Build</span><strong>Contribute on GitHub</strong><p>Code, data packages, the course and this website are open.</p></a>
</div>
```

:::
:::

::: {.owd-band}
::: {.owd-band-inner}

## Stay in touch

<form action="https://buttondown.com/api/emails/embed-subscribe/openwashdata" method="post" class="owd-subscribe">
<label for="owd-email-footer">Subscribe to the monthly newsletter</label>
<input type="email" name="email" id="owd-email-footer" placeholder="your email address" required>
<input type="submit" value="Subscribe">
</form>

:::
:::
