---
pagetitle: "openwashdata"
page-layout: custom
section-divs: false
toc: false
link-external-icon: false
link-external-newwindow: false
execute:
  echo: false
  warning: false
  message: false
listing:
  - id: latest-posts
    contents: pages/blog/posts/*/index.qmd
    type: default
    max-items: 3
    sort: "date desc"
    fields: [date, title]
    sort-ui: false
    filter-ui: false
    categories: false
---

```{r}
#| include: false
library(dplyr)
data <- readr::read_csv(
  "pages/gallery/data/data_data/tbl-01-openwashdata-datasets.csv",
  show_col_types = FALSE
)
pub <- data |> filter(status == "published")
n_pub <- nrow(pub)
n_loc <- n_distinct(pub$location)
first_year <- min(as.integer(format(pub$date_published, "%Y")), na.rm = TRUE)
n_years <- as.integer(format(Sys.Date(), "%Y")) - first_year + 1
```

::: {.owd-hero-a}
::: {.owd-section}

# Open data for water, sanitation and hygiene

<p class="owd-lead">openwashdata is an active global community that applies the FAIR principles to data from the water, sanitation and hygiene (WASH) sector, and teaches the skills to do it.</p>

::: {.owd-actions}
[Explore the data](pages/gallery/data/index.qmd){.btn .btn-primary}
[Join the course](pages/academy/index.qmd){.btn .btn-outline}
:::

::: {#subscribe}
<form action="https://buttondown.com/api/emails/embed-subscribe/openwashdata" method="post" class="owd-subscribe">
<label for="owd-email-hero">One email a month: a dataset, a story, and the community's news.</label>
<input type="email" name="email" id="owd-email-hero" placeholder="your email address" required>
<input type="submit" value="Subscribe">
</form>
:::

:::
:::

::: {.owd-tint}
::: {.owd-section}
::: {.owd-numbers}
::: {.owd-number-item}
<div class="owd-number">`r n_pub`</div>
<div class="owd-number-label">published data packages</div>
:::
::: {.owd-number-item}
<div class="owd-number">`r n_loc`</div>
<div class="owd-number-label">countries and regions</div>
:::
::: {.owd-number-item}
<div class="owd-number">`r n_years`</div>
<div class="owd-number-label">years of publishing, since `r first_year`</div>
:::
:::
:::
:::

::: {.owd-section}

## Dataset of the month

::: {.owd-dotm}
::: {.owd-dotm-text}
**solidwastekampala** holds the quantity and composition of domestic solid waste from 103 households in Kampala, Uganda, weighed over one week in three parishes of low, middle and high income by researchers at Makerere University. Waste per person rises with income, from 0.43 to 0.98 kg per day, and food makes up three quarters of the mass.

[Open solidwastekampala](https://openwashdata.github.io/solidwastekampala){.owd-more}
:::

::: {.owd-dotm-chart}

```{r}
#| label: dataset-of-the-month
#| echo: true
#| code-fold: true
#| code-summary: "Show the R code behind this chart"
#| fig-width: 7
#| fig-height: 4.2
#| fig-alt: "Stacked bar chart of the share of household waste mass by category for the low, middle and high income parishes in Kampala. Food makes up 75 to 82 percent in each parish. Garden and wood, plastics and polythene, other waste, paper and textiles, and glass and metals share the rest."
# pak::pak("openwashdata/solidwastekampala")
library(solidwastekampala)
library(dplyr)
library(tidyr)
library(ggplot2)

groups <- c(
  waste_food_kg = "Food",
  waste_garden_kg = "Garden and wood",
  waste_wood_kg = "Garden and wood",
  waste_polythene_kg = "Plastics and polythene",
  waste_plastics_kg = "Plastics and polythene",
  waste_paper_kg = "Paper and textiles",
  waste_textiles_kg = "Paper and textiles",
  waste_glass_kg = "Glass and metals",
  waste_metals_kg = "Glass and metals",
  waste_other_kg = "Other"
)

shares <- solidwastekampala |>
  select(income_level, parish, all_of(names(groups))) |>
  pivot_longer(starts_with("waste_"), names_to = "category", values_to = "kg") |>
  mutate(group = unname(groups[category])) |>
  summarise(kg = sum(kg, na.rm = TRUE), .by = c(income_level, parish, group)) |>
  mutate(share = kg / sum(kg), .by = income_level)

group_order <- shares |>
  summarise(kg = sum(kg), .by = group) |>
  arrange(desc(kg)) |>
  pull(group)

parish_labels <- shares |>
  distinct(income_level, parish) |>
  arrange(income_level) |>
  mutate(label = paste0(tools::toTitleCase(as.character(income_level)), " income\n(", parish, ")"))

shares <- shares |>
  left_join(parish_labels, by = c("income_level", "parish")) |>
  mutate(
    group = factor(group, levels = rev(group_order)),
    label = factor(label, levels = parish_labels$label)
  )

# chart-1 to chart-6 from openwashdata/brand
chart_colours <- c("#8a3a8a", "#b04208", "#3b5bdb", "#1f8a5f", "#b8860b", "#c2477a")

ggplot(shares, aes(x = label, y = share, fill = group)) +
  geom_col(width = 0.62) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = chart_colours, breaks = group_order) +
  labs(
    title = "Food makes up most of the household waste in Kampala",
    subtitle = "Share of waste mass by category, 103 households in three parishes, one week",
    x = NULL, y = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold", colour = "#1e1e1e"),
    plot.subtitle = element_text(colour = "#5c5c5c"),
    legend.position = "right",
    legend.key.size = unit(0.9, "lines")
  )
```

:::
:::

:::

::: {.owd-tint}
::: {.owd-section}

## Learn

<p class="owd-lead">ds4owd, our free online course, teaches WASH professionals data science in R. Two cohorts have graduated so far.</p>

<p>We are currently preparing the sign-up forms for the third iteration, which starts around September 2027. In the meantime, you can browse the course website of the second cohort from 2025.</p>

[Course website of the 2025 cohort](https://ds4owd-002.github.io/website/){.owd-more}
[About the course](pages/academy/index.qmd){.owd-more}

:::
:::

::: {.owd-section}

## Read

<p class="owd-lead">Stories, tutorials and opinions from and for the community.</p>

::: {#latest-posts}
:::

[All posts](pages/blog/index.qmd){.owd-more}

:::

::: {.owd-tint}
::: {.owd-section}

## Join

```{=html}
<div class="owd-cards">
<a class="owd-card" href="pages/get-started/chat/index.qmd"><span class="owd-card-kicker">Talk</span><strong>Chat with us</strong><p>The community room on Matrix, open to everyone.</p></a>
<a class="owd-card" href="pages/blog/posts/2024-05-17-data-publication-1/index.qmd"><span class="owd-card-kicker">Share</span><strong>Publish your data</strong><p>We turn your WASH dataset into a documented, citable data package with you.</p></a>
<a class="owd-card" href="https://github.com/openwashdata"><span class="owd-card-kicker">Build</span><strong>Contribute on GitHub</strong><p>Code, data packages, the course and this website are open.</p></a>
</div>
```

:::
:::

::: {.owd-band}
::: {.owd-band-inner}

## Stay in touch

<form action="https://buttondown.com/api/emails/embed-subscribe/openwashdata" method="post" class="owd-subscribe">
<label for="owd-email-footer">Subscribe to the monthly newsletter</label>
<input type="email" name="email" id="owd-email-footer" placeholder="your email address" required>
<input type="submit" value="Subscribe">
</form>

:::
:::
