---
pagetitle: "openwashdata"
page-layout: custom
section-divs: false
toc: false
link-external-icon: false
link-external-newwindow: false
execute:
  echo: false
  warning: false
  message: false
listing:
  - id: latest-posts
    contents: pages/blog/posts/*/index.qmd
    type: default
    max-items: 3
    sort: "date desc"
    fields: [date, title, author]
    sort-ui: false
    filter-ui: false
    categories: false
---

```{r}
#| include: false
library(dplyr)
# Same source as the data catalogue: the weekly owdata harvest of the
# organisation (openwashdata/owdata#11).
data <- readr::read_csv(
  "https://raw.githubusercontent.com/openwashdata/owdata/main/inst/extdata/owd_packages.csv",
  show_col_types = FALSE
)
n_pub <- nrow(data)
n_doi <- sum(data$published, na.rm = TRUE)
n_ds <- sum(data$n_datasets, na.rm = TRUE)
first_year <- min(as.integer(format(data$date, "%Y")), na.rm = TRUE)
n_years <- as.integer(format(Sys.Date(), "%Y")) - first_year + 1
```

::: {.owd-hero-a}
::: {.owd-section}

# Open data for water, sanitation and hygiene

<p class="owd-lead">openwashdata is an active global community that applies the FAIR principles to data from the water, sanitation and hygiene (WASH) sector, and teaches the skills to do it.</p>

::: {.owd-actions}
[Explore the data](pages/gallery/data/index.qmd){.btn .btn-primary}
[Read our stories](pages/blog/index.qmd){.btn .btn-outline}
:::

::: {#subscribe}
<form action="https://buttondown.com/api/emails/embed-subscribe/openwashdata" method="post" class="owd-subscribe">
<label for="owd-email-hero">One email a month: a dataset, a story, and the community's news.</label>
<input type="email" name="email" id="owd-email-hero" placeholder="your email address" required>
<input type="submit" value="Subscribe">
</form>
:::

:::
:::

::: {.owd-tint}
::: {.owd-section}
::: {.owd-numbers}
::: {.owd-number-item}
<div class="owd-number">`r n_pub`</div>
<div class="owd-number-label">open data packages, `r n_doi` with a DOI</div>
:::
::: {.owd-number-item}
<div class="owd-number">`r n_ds`</div>
<div class="owd-number-label">datasets inside them</div>
:::
::: {.owd-number-item}
<div class="owd-number">`r n_years`</div>
<div class="owd-number-label">years of publishing, since `r first_year`</div>
:::
:::
:::
:::

::: {.owd-section}

## Dataset of the month

::: {.owd-dotm}
::: {.owd-dotm-text}
**solidwastekampala** holds the quantity and composition of domestic solid waste from 103 households in Kampala, Uganda, weighed over one week in three parishes of low, middle and high income by researchers at Makerere University. Waste per person rises with income, from 0.43 kg per day on average in the low income parish to 0.98 in the high income parish, and food makes up three quarters of the mass.

[Open solidwastekampala](https://openwashdata.github.io/solidwastekampala){.owd-more}
:::

::: {.owd-dotm-chart}

<p class="owd-chart-title">Waste per person rises with income</p>
<p class="owd-chart-note">One dot per household: kg of waste per person per day, with the median of each income level. <span class="owd-chart-hint">Hover over a dot to see that household.</span></p>

```{r}
#| label: dataset-of-the-month-data
#| echo: false
# pak::pak("openwashdata/solidwastekampala")
library(solidwastekampala)
library(dplyr)

households <- solidwastekampala |>
  transmute(
    household = sub("_", " ", household_id),
    income_level,
    label = paste(tools::toTitleCase(as.character(income_level)), "income"),
    parish,
    occupants,
    per_capita = round(waste_per_capita_kg, 2),
    per_day = round(waste_per_day_kg, 1)
  ) |>
  arrange(income_level) |>
  mutate(income_level = as.character(income_level))

medians <- households |>
  summarise(
    median = round(median(per_capita), 2),
    n = n(),
    .by = c(income_level, label, parish)
  )

ojs_define(
  households_data = households,
  medians_data = medians,
  label_order = unique(households$label)
)
```

```{ojs}
//| echo: false
households = transpose(households_data)
medians = transpose(medians_data)

// width of the chart column, so the text keeps its size on every screen
chartWidth = Generators.observe((notify) => {
  const el = document.querySelector(".owd-dotm-chart");
  const report = () => notify(Math.max(280, el.clientWidth));
  const observer = new ResizeObserver(report);
  observer.observe(el);
  report();
  return () => observer.disconnect();
})

// chart-1 and chart-2 from openwashdata/brand
palette = ["#8a3a8a", "#b04208"]

Plot.plot({
  width: chartWidth,
  height: 310,
  marginLeft: 10,
  marginRight: 10,
  marginTop: 6,
  marginBottom: 40,
  style: {fontFamily: "Atkinson Hyperlegible Next, sans-serif", fontSize: 13, background: "transparent"},
  ariaLabel: "Dot plot of waste per person per day for 103 households in Kampala, one row per income level",
  ariaDescription: "The median rises from 0.32 kg in the low income parish to 0.57 in the middle income parish and 0.93 in the high income parish. Households spread from near zero to about 1.9 kg.",
  fy: {domain: label_order, axis: null, padding: 0.25},
  x: {label: "kg of waste per person per day", labelAnchor: "center", domain: [0, 2], grid: true, ticks: 5},
  marks: [
    Plot.dot(households, Plot.dodgeY("middle", {
      fy: "label",
      x: "per_capita",
      r: 3.6,
      fill: palette[0],
      fillOpacity: 0.75,
      stroke: "#ffffff",
      strokeWidth: 0.6,
      tip: true,
      title: (d) => `${d.household}\n${d.per_capita} kg per person per day\n${d.occupants} people, ${d.per_day} kg per day\n${d.label} (${d.parish})`
    })),
    Plot.ruleX(medians, {fy: "label", x: "median", stroke: palette[1], strokeWidth: 2}),
    Plot.text(medians, {
      fy: "label",
      x: "median",
      text: (d) => `median ${d.median} kg`,
      frameAnchor: "bottom",
      textAnchor: "start",
      dx: 5,
      dy: -1,
      fill: palette[1],
      fontWeight: 600
    }),
    // row titles last, with a halo so the median line does not cut through them
    Plot.text(medians, {
      fy: "label",
      text: (d) => `${d.label} (${d.parish}), ${d.n} households`,
      frameAnchor: "top-left",
      dx: 2,
      dy: -2,
      fontWeight: 700,
      fill: "#1e1e1e",
      stroke: "#f4f4f4",
      strokeWidth: 5
    })
  ]
})
```

:::
:::

:::

::: {.owd-tint}
::: {.owd-section}

## Learn

<p class="owd-lead">Data Science for openwashdata (ds4owd), our free online course, teaches WASH professionals data science in R. Two cohorts have graduated so far. A third cohort will start in September 2027. We are currently preparing the sign-up forms for the third iteration. In the meantime, you can browse the course website of the second cohort from 2025.</p>

[Course website of the 2025 cohort](https://ds4owd-002.github.io/website/){.owd-more}
[Projects of the 2026 graduates](pages/academy/graduates-2026/index.qmd){.owd-more}
[About the course](pages/academy/index.qmd){.owd-more}

:::
:::

::: {.owd-section}

## Read

<p class="owd-lead">Stories, tutorials and opinions from and for the community.</p>

::: {#latest-posts}
:::

[All posts](pages/blog/index.qmd){.owd-more}

:::

::: {.owd-tint}
::: {.owd-section}

## Join

```{=html}
<div class="owd-cards">
<a class="owd-card" href="pages/get-started/chat/index.qmd"><span class="owd-card-kicker">Talk</span><strong>Chat with us</strong><p>The community room on Matrix, open to everyone.</p></a>
<a class="owd-card" href="pages/blog/posts/2024-05-17-data-publication-1/index.qmd"><span class="owd-card-kicker">Share</span><strong>Publish your data</strong><p>We turn your WASH dataset into a documented, citable data package with you.</p></a>
<a class="owd-card" href="https://github.com/openwashdata"><span class="owd-card-kicker">Build</span><strong>Contribute on GitHub</strong><p>Code, data packages, the course and this website are open.</p></a>
</div>
```

:::
:::

::: {.owd-band}
::: {.owd-band-inner}

## Stay in touch

<form action="https://buttondown.com/api/emails/embed-subscribe/openwashdata" method="post" class="owd-subscribe">
<label for="owd-email-footer">Subscribe to the monthly newsletter</label>
<input type="email" name="email" id="owd-email-footer" placeholder="your email address" required>
<input type="submit" value="Subscribe">
</form>

:::
:::
