---
title: "Workshop"
subtitle: "GHE & WASHWeb"
author: 
  - name: "Lars Schöbitz"
    url: https://openwashdata.org/about/lars/
    affiliation: Global Health Engineering, ETH Zurich
    affiliation_url: https://ghe.ethz.ch/
date: 2024-04-09
date-format: long
format: 
  revealjs:
    theme: "../../../../style/slides.scss"
    transition: fade
    slide-number: true
    preview-links: auto
editor: source
editor_options: 
  chunk_output_type: console
bibliography: ["../datasets.bib", "../cu-wash-symposium-2023-openwashdata/references.bib"]
---

```{r}

library(washopenresearch)
library(tidyverse)
library(ggthemes)
library(ggtext)
library(gt)
library(countdown)

# set ggplot theme --------------------------------------------------------

ggplot2::theme_set(
    # Pick a starting theme
    ggplot2::theme_gray(base_size = 14) +
        # Add your favourite elements
        ggplot2::theme(
            axis.line = element_line(),
            panel.background = element_rect(fill = "white"),
            panel.grid.major = element_line("grey95", linewidth = 0.25),
            legend.key  = element_rect(fill = NA) 
        )
)

```

# Getting to know each other - Speed interview activity 

## {{< fa circle-question >}} {.center-align auto-animate="true"}

::: {style="margin-top: 50px; font-size: 1.5em"}
Coffee or Tea?
:::

## {{< fa circle-question >}} {.center-align auto-animate="true"}

::: {style="margin-top: 50px; font-size: 1.5em"}
Twitter or LinkedIn?
:::

## {{< fa circle-question >}} {.center-align auto-animate="true"}

::: {style="margin-top: 50px; font-size: 1.5em"}
Do you have a personal role model and if so who is it?
:::

## {{< fa circle-question >}} {.center-align auto-animate="true"}

::: {style="margin-top: 50px; font-size: 1.5em"}
What is your favorite place to travel? Why?
:::

## {{< fa circle-question >}} {.center-align auto-animate="true"}

::: {style="margin-top: 50px; font-size: 1.5em"}
What do you think is the best invention of our existence?
:::

## {{< fa circle-question >}} {.center-align auto-animate="true"}

::: {style="margin-top: 50px; font-size: 1.5em"}
Something that nobody teaches you, but that everyone should learn.
:::

# Getting to know each other - Round of personal introductions

##  {{< fa handshake-simple >}} {auto-animate="true"}

::: task

- [Past]{.highlight-peach}: How did you end up caring about data and WASH?
- [Present]{.highlight-peach}: What are you working on right now?
- [Future]{.highlight-peach}: What are you hoping to do in the future?
- [You]{.highlight-peach}: Add a personal note about yourself, if you like. 

::: hand
Think & take notes for 2 minutes, then share for 1 minute
:::
:::

```{r}
#| echo: false

countdown(minutes = 2)
```

# Getting to know each other - Brief overview of GHE

# Getting to know each other - Introduction to WASHWeb vision & working groups