---
pagetitle: "Data"
description: |
    A searchable catalogue of the datasets published through the openwashdata
    community, and of the variables inside them.
editor_options:
  chunk_output_type: console
execute:
  echo: false
  warning: false
---

```{r}
library(DT)
library(dplyr)
```

```{r}
# The catalogue comes from the owdata harvest, which reads every repository in
# the organisation once a week. Until washr 1.1.0 this page read a Google Sheet
# that nothing writes to any more (openwashdata/owdata#11).
owd_base <- "https://raw.githubusercontent.com/openwashdata/owdata/main/inst/extdata/"

packages <- readr::read_csv(paste0(owd_base, "owd_packages.csv"), show_col_types = FALSE)
variables <- readr::read_csv(paste0(owd_base, "owd_variables.csv"), show_col_types = FALSE)

readr::write_csv(packages, here::here("pages/gallery/data/data_data/owd_packages.csv"))
readr::write_csv(variables, here::here("pages/gallery/data/data_data/owd_variables.csv"))

n_pkg <- nrow(packages)
n_doi <- sum(packages$published, na.rm = TRUE)
```

## Data

The table below lists every R data package in the openwashdata organisation.
Each one comes with its own website and documentation, so the data can be
reused and cited.

A package has a **DOI** once it is deposited on Zenodo, which is what makes it
citable and permanently archived. Of the `r n_pkg` packages listed here,
`r n_doi` have one. The rest are published as packages and usable today, and
are waiting on a deposit.

:::{.column-page}
```{r}
tbl <- packages |>
  mutate(
    Package = paste0("[", pkg_name, "](", url_docs, ")"),
    DOI = ifelse(
      is.na(doi),
      "",
      paste0("[", doi, "](https://doi.org/", doi, ")")
    ),
    Updated = format(date, "%d %B %Y")
  ) |>
  select(
    Package,
    Description = description,
    Datasets = n_datasets,
    DOI,
    Updated
  ) |>
  arrange(Package)

datatable(
  tbl,
  escape = FALSE,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 25,
    lengthMenu = c(10, 25, 50, 100),
    columnDefs = list(list(width = "45%", targets = 1))
  )
)
```
:::

## Variables

Every variable documented across the packages with a DOI. Search here to find
which dataset holds the measurement you need.

:::{.column-page}
```{r}
var_tbl <- variables |>
  left_join(select(packages, pkg_name, url_docs), by = "pkg_name") |>
  mutate(Package = paste0("[", pkg_name, "](", url_docs, ")")) |>
  select(
    Package,
    Dataset = dataset_name,
    Variable = variable_name,
    Type = variable_type,
    Description = description
  ) |>
  arrange(Package, Dataset, Variable)

datatable(
  var_tbl,
  escape = FALSE,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 25,
    lengthMenu = c(10, 25, 50, 100)
  )
)
```
:::

```{r}
#| output: asis
cat(
  "\n*Catalogue harvested from the openwashdata organisation by ",
  "[owdata](https://github.com/openwashdata/owdata). ",
  "Most recent package update: ",
  format(max(packages$date, na.rm = TRUE), "%d %B %Y"), ".*\n",
  sep = ""
)
```
