---
title: "openwashdata & WASHWeb partnership workshop"
description: |
  This workshop is hosted by the Global Health Engineering group at ETH Zurich. 
  It's our first time to meet in person and the goals are to get to know each other,
  discuss on items of collaboration and define task responsibilities, and 
  actively work on material for collaborative events.
bibliography: ../../../references.bib
execute:
    echo: false
    warning: false
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r}
library(tidyverse)
```

```{r}
options(knitr.kable.NA = '')
```


## Participants

- Merel Laauwen, [WASHWeb](https://washweb.org/)
- Nick Dickinson, [WASHWeb](https://washweb.org/)
- Elizabeth Tilley, [Global Health Engineering, ETH Zurich](https://ghe.ethz.ch/)
- Colin Walder, [Global Health Engineering, ETH Zurich](https://ghe.ethz.ch/)
- Lars Schöbitz, [Global Health Engineering, ETH Zurich](https://ghe.ethz.ch/)


## 📆 Event Details

- Dates: 08 April & 09 April 2024
- Location: Zurich, Switzerland

## 📝 Agenda

### Day 1

```{r}
#| tbl-colwidths: [50, 35, 15, 30]
#| column: page
#| eval: true

agenda <- readr::read_csv(here::here("pages/events/2024-04-09-workshop-washweb/data/tbl-09-workshop-ghe-washweb.csv"))

agenda |> 
    filter(day == 1) |> 
    dplyr::select(time, title, lead) |> 
    dplyr::rename_with(~stringr::str_to_title(.x), everything()) |> 
    knitr::kable()
    
```

### Day 2

```{r}
#| tbl-colwidths: [50, 35, 15, 30]
#| column: page
#| eval: true

agenda |> 
    filter(day == 2) |> 
    dplyr::select(time, title, lead) |> 
    dplyr::rename_with(~stringr::str_to_title(.x), everything()) |> 
    knitr::kable()
```


## Resources

1. openwashdata vision presentation: <https://openwashdata.org/pages/gallery/slides/washweb-webinar/#/title-slide>
2. openjmp funded proposal: <https://github.com/openwashdata-dev/prop-eth-ord-explore-02-openwashdata/blob/master/proposal.md>
3. openjmp slides: <https://openwashdata.org/pages/gallery/slides/cu-wash-symposium-2023-openjmp/#/title-slide>
4. jmpwashdata R package: <https://github.com/WASHNote/jmpwashdata>

